---
title: "lab_fundamentals_1"
author: "Evan Krause"
date: "2022-09-07"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1.the first output is a vector, the second is a string
```{r}
c_1 <-  c(1,2,3)
c_2 <- "c(1,2,3)"
```

2.c_1 is a variable, there is no input/output just an assigned value
3.c_2 is a variable for the same reason as above
4.C_1 AND c_2 have different values because they're different classes of object

```{r}
my_vec <- 1:6
mat_1 <- matrix(my_vec, nrow = 3)
```

5.[3,2] 3 rows two columns

6.
```{r}
mat_1[3,1]
```

```{r}
mat_2 <- matrix(my_vec, nrow = 2, ncol = 3)
mat_3 <- matrix(my_vec, nrow = 3, ncol = 2)
```

7. mat_2 <- matrix(my_vec, nrow = 2, ncol = 3)
8.mat_3 <- matrix(my_vec, nrow = 3, ncol = 2)
9.columns

10.
```{r}
mat_4 <- matrix(my_vec, )
```
11.

```{r}
my_list_1 <- list(5.2, "five point two", 1:5)
```

12.
value,
value,
null,
null,
null,
null,
error,
null
13.none, the second element of the list was never called
14.3,4,5,6 and 8 were null because they were invalid elements




